Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-11-17 paper/main.tex
update results on paper

In the case of the QAP and PFSP, we consider the same instances as
\citet{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, i.e.,
\texttt{nug12}, \texttt{nug30}, \texttt{tho30} and \texttt{kra32} for the QAP,
with $n$ (the permutation length) equal to the number in the instance name, and \texttt{reC05},
\texttt{reC13}, \texttt{reC19}, \texttt{reC31} for the PFSP, with
$n \in \{20, 20, 30, 50\}$, respectively.

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra32}. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet CEGO appears to get stuck around
evaluation 250, and UMM overtakes it around evaluation 300. We believe the
reason is the same as before, that is, the CEGO model is not able to adequately
reflect the ruggedness of the actual landscape and, hence, the best solutions
found by the GA searching the model are not improved solutions for the actual
problem. Interestingly, there are well-known differences between
\texttt{kra32}, on one hand, and \texttt{nug12}, \texttt{nug30} and
\texttt{tho32}, on the other~\citep{AnsBriGou2002qap}. However, we cannot at
this moment point out the root cause of this different behavior.

@@

In the LOP, differences are mostly in favour of UMM, due to the time limit of
144 hours, which resulted in CEGO typically achieving less than 200
evaluations. In contrast, UMM was always able to reach 400 evaluations in less
than 16 hours. If UMM was allowed to reach the 144 hours limit, the differences
would be even stronger.

In the case of the PFSP, the differences are minor and/or not statistically
significant, with the qualification that both algorithms appear to be
performing a random search. Similar observations can be made for the QAP, where
only on the smallest instance \texttt{nug12}, there is a significant different
in favour of CEGO.

@@

 \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean of the difference between the fitness of UMM minus the one of CEGO. Mean runtime is measured in hours. If a run of  CEGO was not able to reach 400 function evaluations (FEs) in 144 hours, it was terminated earlier. This was the case for all LOP instances shown.\manuel{}{Don't we have results for rec31?}\label{tab:results}}

@@

    LOP     & N-t1d100.01      & 78696.6                 & (811.4)    & 76119.6   & (915.4)                & $[$1763.5,    & 3390.5$]$     & 111.8& 72.5 & 3.2  \\
     & N-t1d100.02      & 79686.5                 & (367.9)    & 76827.7   & (891.0)                & $[$2194.5,    & 3523.1$]$     & 110.7& 72.9 & 3.2  \\
     & N-t1d150.01      & 187605.5                & (1906.9)   & 179508.3  & (1647.9)               & $[$6420.3,    & 9774.1$]$     & 73.9 & 72.7 & 6.5  \\
     & N-t1d150.02      & 184320.9                & (973.5)    & 177592.8  & (1196.2)               & $[$5700.5,    & 7755.7$]$     & 74.1 & 73.2 & 6.5  \\
     & N-t1d200.01      & 335391.1                & (2289.0)   & 323513.6  & (1371.2)               & $[$10076.1,   & 13678.9$]$    & 55.6 & 73.2 & 10.7 \\
     & N-t1d200.02      & 333149.5                & (2497.2)   & 319675.8  & (2242.6)               & $[$11242.0,   & 15705.4$]$    & 56.5 & 73.7 & 10.6 \\
     & N-t2d150.01      & 24682.7                 & (976.6)    & 15297.7   & (563.5)                & $[$8622.2,    & 10147.8$]$    & 74.6 & 72.9 & 7.0  \\
     & N-t2d150.02      & 24250.8                 & (1354.9)   & 14724.9   & (667.5)                & $[$8495.1,    & 10556.7$]$    & 74.8 & 72.8 & 7.1  \\
     & N-t2d200.01      & 57098.6                 & (4452.8)   & 33465.2   & (2001.0)               & $[$20284.5,   & 26982.3$]$    & 56.3 & 73.3 & 11.2 \\
     & N-t2d200.02      & 54895.6                 & (4589.7)   & 35901.8   & (2077.5)               & $[$15539.1,   & 22448.5$]$    & 55.6 & 73.5 & 11.2 \\
     & N-p50-01         & 21431.2                 & (299.5)    & 21027.6   & (711.8)                & $[$-128.1,    & 935.3$]$      & 219.9& 72.3 & 0.9  \\
     & N-p50-02         & 22021.1                 & (275.6)    & 21613.6   & (596.4)                & $[$-42.5,     & 857.5$]$      & 217.3& 72.3 & 0.9  \\
     & N-atp111         & 693.1                   & (10.3)     & 594.0     & (15.4)                 & $[$86.7,      & 111.5$]$      & 98.7 & 72.5 & 4.1  \\
     & N-atp134         & 880.5                   & (18.6)     & 756.3     & (21.3)                 & $[$105.4,     & 143.0$]$      & 83.0 & 72.8 & 5.7  \\
     & N-be75eec\_150   & 1666193.1               & (72310.8)  & 1415208.5 & (58593.5)              & $[$188960.6,  & 313008.6$]$   & 74.0 & 72.9 & 6.9  \\
     & N-be75eec\_250   & 5214413.0               & (166701.4) & 4431271.0 & (109756.7)             & $[$649040.8,  & 917243.2$]$   & 44.9 & 73.7 & 15.5 \\
     & N-be75np\_150    & 3742444.0               & (88500.2)  & 3204878.3 & (80704.1)              & $[$457944.2,  & 617187.2$]$   & 74.2 & 72.9 & 6.9  \\
     & N-be75np\_250    & 10887427.8              & (159642.5) & 9544754.9 & (260523.9)             & $[$1136635.8, & 1548710.0$]$  & 45.3 & 74.0 & 15.5 \\\midrule
   PFSP     & rec05        & 1317.9                  & (25.1)     & 1330.7    & (17.9)                 & $[$-33.4,     & 7.8$]$        & 400.0& 38.9 & 0.1  \\
       & rec13        & 2128.2                  & (25.0)     & 2158.7    & (29.5)                 & $[$-56.2,     & -4.8$]$       & 400.0& 38.9 & 0.1  \\
       & rec19        & 2398.6                  & (19.8)     & 2405.1    & (12.6)                 & $[$-22.3,     & 9.3$]$        & 400.0& 82.1 & 0.3  \\\midrule
    QAP  & nug12        & 624.2                   & (26.3)     & 660.4     & (17.3)                 & $[$-57.3,     & -15.1$]$      & 400.0& 14.8 & 0.0  \\
         & nug30        & 7486.8                  & (40.4)     & 7474.0    & (83.9)                 & $[$-50.8,     & 76.4$]$       & 400.0& 84.3 & 0.3  \\
        & tho30        & 192416.0                & (2584.2)   & 191482.0  & (4419.7)               & $[$-2527.0,   & 4395.0$]$     & 400.0& 85.1 & 0.3  \\
         & kra32        & 115907.0                & (4663.4)   & 116320.0  & (4673.5)               & $[$-4799.3,   & 3973.3$]$     & 400.0& 95.8 & 0.3  \\

>>>>

In the case of the QAP and PFSP, we consider the same instances as
\citet{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, i.e., \texttt{nug12},
\texttt{nug30}, \texttt{tho30} and \texttt{kra32} for the QAP, with $n$ (the
permutation length) equal to the number in the instance name, and
\texttt{reC05}, \texttt{reC13} and \texttt{reC19} for the PFSP, with
$n \in \{20, 20, 30\}$, respectively. For reasons that will become obvious
later, we also consider QAP instances \texttt{kra30a} and \texttt{kra30b}.

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra} instances. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet CEGO appears to get stuck around
evaluation 250, and UMM overtakes it around evaluation 300. We believe the
reason is the same as before, that is, the CEGO model is not able to adequately
reflect the ruggedness of the actual landscape and, hence, the best solutions
found by the GA searching the model are not improved solutions for the actual
problem. Interestingly, there are well-known differences between
\texttt{kra30a}, \texttt{kra30b}, \texttt{kra32}, on one hand, and \texttt{nug12}, \texttt{nug30} and
\texttt{tho32}, on the other~\citep{AnsBriGou2002qap}. However, we cannot at
this moment point out the root cause of this different behavior.

@@

In the LOP, differences are mostly in favour of UMM, due to the time limit of
144 hours, which resulted in CEGO typically achieving less than 200
evaluations. In contrast, UMM was always able to reach 400 evaluations in less
than 12.5 hours. If UMM was allowed to reach the 144 hours limit, the differences
would be even stronger.

In the case of the PFSP, the differences are minor and/or not statistically
significant, with the qualification that both algorithms appear to be
performing a random search. Similar observations can be made for the QAP, where
only on the smallest instance \texttt{nug12}, there is a significant difference
in favour of CEGO.

@@

 \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean of the difference between the fitness of UMM minus the one of CEGO. Mean runtime is measured in hours. If a run of  CEGO was not able to reach 400 function evaluations (FEs) in 144 hours, it was terminated earlier. This was the case for all LOP instances shown.\label{tab:results}}

@@

LOP
& N-p50-01       & 21034.3             & (183.9)            & 21161.8            & (657.4)           & [-606.0,             & 351.0]              & 305.8        & 144.6 & 0.5  \\
& N-p50-02       & 21767.3             & (254.9)            & 21746.6            & (645.4)           & [-458.6,             & 500.0]              & 315.1        & 144.5 & 0.5  \\
& N-t1d100.01    & 78012.0             & (432.1)            & 76499.3            & (1099.7)          & [696.4,              & 2329.0]             & 158.3        & 145.1 & 1.9  \\
& N-t1d100.02    & 78834.0             & (477.8)            & 76860.7            & (864.0)           & [1303.8,             & 2642.8]             & 156.0        & 144.7 & 1.8  \\
& N-atp111       & 668.5               & (9.4)              & 596.6              & (17.0)            & [58.8,               & 85.0]               & 137.9        & 145.1 & 2.3  \\
& N-atp134       & 856.4               & (12.2)             & 760.1              & (23.4)            & [78.4,               & 114.2]              & 116.5        & 145.0 & 3.3  \\
& N-t1d150.01    & 186969.0            & (2094.6)           & 179615.6           & (1731.5)          & [5543.3,             & 9163.5]             & 103.9        & 145.7 & 4.4  \\
& N-t1d150.02    & 183414.7            & (779.2)            & 177524.2           & (1165.6)          & [4949.2,             & 6831.8]             & 103.6        & 145.3 & 4.6  \\
& N-t2d150.01    & 22207.3             & (1667.3)           & 15136.9            & (745.9)           & [5817.1,             & 8323.7]             & 104.5        & 145.5 & 4.5  \\
& N-t2d150.02    & 22036.0             & (744.7)            & 14718.4            & (677.5)           & [6648.3,             & 7986.9]             & 103.1        & 145.0 & 4.5  \\
& N-t1d200.01    & 333804.1            & (1191.9)           & 323501.4           & (1385.7)          & [9086.4,             & 11519.0]            & 77.7         & 145.4 & 7.6  \\
& N-t1d200.02    & 331730.9            & (2819.7)           & 319786.2           & (2265.6)          & [9533.6,             & 14355.8]            & 78.2         & 145.9 & 7.1  \\
& N-t2d200.01    & 54043.6             & (1477.6)           & 33389.0            & (1984.0)          & [19001.4,            & 22307.8]            & 78.2         & 146.7 & 7.4  \\
& N-t2d200.02    & 52787.0             & (4194.2)           & 35589.0            & (2319.6)          & [13948.0,            & 20448.0]            & 78.9         & 145.7 & 7.5  \\
& N-be75np\_150  & 3669753.9           & (74292.3)          & 3216907.1          & (72239.3)         & [383998.5,           & 521695.1]           & 105.2        & 145.5 & 4.4  \\
& N-be75np\_250  & 10818828.5          & (127722.3)         & 9582261.7          & (258176.8)        & [1040022.9,          & 1433110.7]          & 62.5         & 146.4 & 12.5 \\
& N-be75eec\_150 & 1592220.8           & (29945.9)          & 1418578.9          & (56557.7)         & [130141.0,           & 217142.8]           & 103.2        & 145.9 & 4.4  \\
& N-be75eec\_250 & 5098734.4           & (172874.0)         & 4430408.5          & (110233.0)        & [530352.4,           & 806299.4]           & 63.0         & 146.7 & 12.3 \\\midrule
PFSP
& rec05          & 1317.9              & (25.1)             & 1332.5             & (15.6)            & [-34.5,              & 5.3]                & 400.0        & 38.6  & 0.1  \\
& rec13          & 2128.2              & (25.0)             & 2163.6             & (28.1)            & [-60.4,              & -10.4]              & 400.0        & 38.6  & 0.1  \\
& rec19          & 2398.6              & (19.8)             & 2405.2             & (12.8)            & [-22.5,              & 9.3]                & 400.0        & 85.2  & 0.2  \\\midrule
QAP
& nug12          & 624.2               & (26.3)             & 661.2              & (17.8)            & [-58.3,              & -15.7]              & 400.0        & 14.2  & 0.0  \\
& kra30b         & 120192.0            & (1778.1)           & 119923.0           & (3459.6)          & [-2379.5,            & 2917.5]             & 400.0        & 85.1  & 0.2  \\
& tho30          & 192416.0            & (2584.2)           & 191586.4           & (4342.0)          & [-2582.9,            & 4242.1]             & 400.0        & 85.6  & 0.2  \\
& kra30a         & 119872.0            & (1746.7)           & 119502.0           & (1827.5)          & [-1309.8,            & 2049.8]             & 400.0        & 86.1  & 0.2  \\
& nug30          & 7486.8              & (40.4)             & 7472.4             & (83.6)            & [-49.0,              & 77.8]               & 400.0        & 86.5  & 0.2  \\
& kra32          & 115907.0            & (4663.4)           & 116898.0           & (4343.1)          & [-5226.3,            & 3244.3]             & 400.0        & 94.0  & 0.2  \\
